
title: "PROJECT 2: Crocodile Data Analysis"
author: "James Ross and Berenice Gonzalez"
subtitle:"CPSC 292 Section 02"
>>>>>>> 72234e6b41277fee8d78485509e94984a99c1807
output: html_document
date: "2025-10-31"
---

## libraries and functions

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(viridis)
library(janitor)
library(grid)
library(tidyverse)
library(dplyr)
```


## Figures

```{r Original Graph Draft(ignore)}
croc <- read.csv("Data/crocodile_dataset.csv")

ages <- c("Hatchling", "Juvenile", "Subadult", "Adult")
ageorder<- factor(croc$Age.Class, levels = ages)


"Observed Length (m)" <- croc$Observed.Length..m.

ggplot(croc, aes(x = Genus, y = ageorder, fill = `Observed Length (m)`)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "darkgreen") +
  xlab("Crocodile Genus")+
  ylab("Age Estimate")+
  ggtitle("Crocodile Observed Lengths per age within Genus") +
  theme_bw()

```


## Final Project 2 Graphics

```{r Project 2}

croc <- read.csv("Data/crocodile_dataset.csv")

ages <- c("Hatchling", "Juvenile", "Subadult", "Adult")
ageorder <- factor(croc$Age.Class, levels = ages)
"Observed Length (m)" <- croc$Observed.Length..m.

for (g in unique(croc$Genus)) {
  avg_length <- mean(croc$Observed.Length..m.[croc$Genus == g], na.rm = TRUE)
  print(paste("Average observed length for", g, ":", round(avg_length, 2), "m"))
}

  if (avg_length > 3) {
    print(paste(g, "are large crocodiles with an average length of", round(avg_length, 2), "m."))
  } else if (avg_length > 2) {
    print(paste(g, "are medium-sized crocodiles with an average length of", round(avg_length, 2), "m."))
  } else {
    print(paste(g, "are relatively small crocodiles with an average length of", round(avg_length, 2), "m."))
  }

ggplot(croc, aes(x = Genus, y = ageorder, fill = `Observed Length (m)`)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "darkgreen") +
  xlab("Crocodile Genus") +
  ylab("Age Estimate") +
  ggtitle("Crocodile Observed Lengths per Age within Genus") +
  theme_bw()

croc$Size.Score <- ifelse(croc$Observed.Length..m. > 3, 3,
                     ifelse(croc$Observed.Length..m. > 2, 2, 1))
Sizes <- croc$Size.Score

ggplot(croc, aes(x = Genus, y = ageorder, fill = Sizes)) +
  geom_tile(color = "black") +
  scale_fill_gradient(low = "tan", high = "darkgreen") +
  xlab("Crocodile Genus") +
  ylab("Age Estimate") +
  ggtitle("Crocodile Observed Sizes per Age within Genus") +
  theme_bw()

```


#SUMMARY

For our project, we worked with a dataset containing information on crocodile genera, age classes, and observed lengths. We organized the age groups, which were the Hatchling, Juvenile, Subadult, Adult age groups and added a loop with three elements representing the three genera in the dataset. Inside the loop, we used a conditional statement to print the average observed length for each type of crocodile, categorizing them as large, medium, or small based on their average size. We then created a ggplot2 heatmap to visualize the observed lengths across different age classes and genera, using a color gradient to highlight size differences. This allowed us to both analyze and visually compare crocodile growth patterns among species. After receiving feedback on our draft, we added two control elements that contribute directly to a second visualization. In this improved figure, we used an ifelse statement to assign a numerical size score (1, 2, or 3) depending on the crocodile’s observed length. We then incorporated this new variable into a second heatmap that again showcases crocodile size per age within each genus. These additions ensure that our control structures directly influence the visualizations, creating clearer and more meaningful results.



Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
