---
title: "PROJECT 2: Crocodile Data Analysis"
author: "James Ross and Berenice Gonzalez"
subtitle:"CPSC 292 Section 02"
output: html_document
date: "2025-10-31"
---

## libraries and functions

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(viridis)
library(janitor)
library(grid)
library(tidyverse)
library(dplyr)
```


## Figures

```{r Original Graph Draft(ignore)}
croc <- read.csv("Data/crocodile_dataset.csv")

ages <- c("Hatchling", "Juvenile", "Subadult", "Adult")
ageorder<- factor(croc$Age.Class, levels = ages)


"Observed Length (m)" <- croc$Observed.Length..m.

ggplot(croc, aes(x = Genus, y = ageorder, fill = `Observed Length (m)`)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "darkgreen") +
  xlab("Crocodile Genus")+
  ylab("Age Estimate")+
  ggtitle("Crocodile Observed Lengths per age within Genus") +
  theme_bw()

```


## Final Project 2 Graphics

```{r Project 2}

croc <- read.csv("Data/crocodile_dataset.csv")

ages <- c("Hatchling", "Juvenile", "Subadult", "Adult")
ageorder <- factor(croc$Age.Class, levels = ages)

for (g in unique(croc$Genus)) {
  avg_length <- mean(croc$Observed.Length..m.[croc$Genus == g], na.rm = TRUE)
  print(paste("Average observed length for", g, ":", round(avg_length, 2), "m"))}

  if (avg_length > 3) {
    print(paste(g, "are large crocodiles with an average length of", round(avg_length, 2), "m."))
  } else if (avg_length > 2) {
    print(paste(g, "are medium-sized crocodiles with an average length of", round(avg_length, 2), "m."))
  } else {
    print(paste(g, "are relatively small crocodiles with an average length of", round(avg_length, 2), "m."))
  }

ggplot(croc, aes(x = Genus, y = ageorder, fill = `Observed Length (m)`)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "darkgreen") +
  xlab("Crocodile Genus") +
  ylab("Age Estimate") +
  ggtitle("Crocodile Observed Lengths per Age within Genus") +
  theme_bw()


```


#SUMMARY

For our project, we worked with a dataset containing information on crocodile genera, age classes, and observed lengths. We organized the age groups, which were the Hatchling, Juvenile, Subadult, Adult age groups and added a loop with three elements representing the three genera in the dataset. Inside the loop, we used a conditional statement to print the average observed length for each type of crocodile, categorizing them as large, medium, or small based on their average size. We then created a ggplot2 heatmap to visualize the observed lengths across different age classes and genera, using a color gradient to highlight size differences. This allowed us to both analyze and visually compare crocodile growth patterns among species.



Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
